An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2021-08-26 latex_files/ch2/2_4_ROMC.tex
small changes

\begin{itemize}
\item The whole acceptable region $\mathcal{S}_i$
  shrinks to a single point $\thetab_i^*$. This simplification adds
  significant apprimation error when then the area $\mathcal{S}_i$ is relatively big.
\item The weight $w_i$ is computed based only at the curvature at the
  point $\thetab_i^*$. Using only local information for estimating the weight can also be misleading. For example, if $g_i$ is almost flat at $\thetab_i^*$, i.e. $\text{det}(\jac_i^{*T}\jac_i^*) \rightarrow 0, then \Rightarrow w_i
  \rightarrow \infty$ dominating the posterior.
\item There is no way to solve the optimisation problem
  $\thetab_i^* = \text{argmin}_{\thetab} \: g_i(\thetab)$ when $g_i$
  is not differentiable.
\end{itemize}

@@

ROMC resolves the failure-modes of OMC mainly by approaching the
acceptance regions $\mathcal{S}_i$ in a more sophisticated way.
Instead of collapsing the regions into single points, it approximates
them with a bounding box. Afterwards, a uniform distribution defined
on the bounding box area, is used as the \emph{proposal distribution}
for importance sampling. Defining as $q_i$ the uniform distribution
defined on the $i-th$ bounding box, weighted sampling is performed by
drawing samples from $q_i$ and assigning the weights as
in~\eqref{eq:sampling}.

@@

The constrution of the proposal regions $q_i$ is performed in two
steps; (a) solving the deterministic optimisation problems as in OMC
and (b) constructing the bounding box regions.

@@

For each set of nuisance variables $\vb_i, i = \{1,2,\ldots,n_1 \}$,
we search for a point $\thetab^*_i$ such that
$d(M_d(\thetab^*_i,\vb_i), \data) < \epsilon$. For notation
convenience, we name the distance function $d(M_d(\thetab,\vb_i), \data)$ as
$d_i(\thetab)$.  Obtaining this point involves solving the following
optimisation problem:

\begin{subequations}
\begin{alignat}{2}
  &\!\min_{\thetab}        &\qquad& d_i(\thetab) = d(M_d(\thetab^*,\vb_i), \data)\label{eq:optProb}\\
  &\text{subject to} & & d_i(\thetab) \leq \epsilon
\end{alignat}
\end{subequations}
%
The optimisation problem can be treated as unconstrained, accepting
the optimal point $\thetab_i^* = \text{argmin}_{\thetab} d_i(\thetab)$
only if $d_i(\thetab_i^*) < \epsilon$. The nature of the generative
model $M_r(\thetab)$, specifies the properties of the objective
function $d_i$. If $d_i$ is differentiable any gradient-based
iterative algorithm can be used for~\ref{eq:optProb}. The gradients
$\nabla_{\thetab} d_i$ can be either provided in closed form or
approximated by finite differences. Alternatively, if $d_i$ is not
differentiable, the Bayesian Optimisation
approach~\cite{Shahriari2016} provides an alternative choice. In this
case, apart from obtaining an optimal $\thetab_i^* $, a surrogate
model $\hat{d}_i$ of the distance $d_i$ is also fitted; the
approximate model can then substitute the actual model in the
following steps.

\subsubsection*{Proposal region computation}

After obtaining $\thetab_i^*$, we initially approximate the acceptance
region $\mathcal{S}_i$ by constructing a bounding box
$\mathcal{\hat{S}}_i$ around it. The bounding must the part of the
acceptance region $\mathcal{S}_i$ \emph{around} $\thetab_i^*$;
formally $\{ \thetab : d_i(\thetab) < \epsilon$,
$d(\thetab, \thetab_i^*) < M \}$. The condition
$d(\thetab, \thetab_i^*) < M$ describes that if
$\mathcal{S}_i := \{ \thetab : d_i(\thetab) < \epsilon \} $ contains
many \emph{disjoint} sets of $\thetab$ that they fulfill
$d_i(\thetab) < \epsilon$, we want the bounding box to match only the
set that contains $\thetab_i^*$. Ideally, we want the bounding box to
be as tight as possible to $\mathcal{S}_i$, to ensure high acceptance
rate in the importance sampling, and big enough for not discarding
valid parts of $\mathcal{S}_i$.

The bounding boxes are built in two steps; firstly, we define their
axes $\mathbf{v}_d$ in the $d$-dimensional space and, then, the limits
are obtained with a line-search method looking for the smallest
$\kappa: d_i(\thetab_i^* + \kappa \vb_d) \geq
\epsilon$\footnote{$-\kappa$ is used as well for the opposite
  direction along the search
  line}. Algorithm~\ref{alg:region_construction} describes the method
in-depth. After limits are found along all directions, the uniform
distribution $q_i$ is defined on the bounding box, and is used as the
proposal region for importance sampling.

>>>>

\begin{itemize}
\item The whole acceptable region $\mathcal{S}_i$
  shrinks to a single point $\thetab_i^*$. This simplification adds
  significant approximation error when then the area $\mathcal{S}_i$ is relatively big.
\item The weight $w_i$ is computed based only at the curvature at the
  point $\thetab_i^*$. Using only local information for estimating the
  weight can also be misleading. For example, if $g_i$ is almost flat
  at $\thetab_i^*$, i.e.
  $\text{det}(\jac_i^{*T}\jac_i^*) \rightarrow 0, then w_i \rightarrow
  \infty$, dominating the posterior.
\item There is no way to solve the optimisation problem
  of~\eqref{eq:omc_opt_prob} when
  $||\Phi(\data)-\Phi(M_d(\thetab, \vb_i))||_2$ is not differentiable.
\end{itemize}

@@

ROMC resolves the failure-modes of OMC, mainly, by approaching the
acceptance regions $\mathcal{S}_i$ in a more sophisticated way;
instead of collapsing the regions into single points, it approximates
them with a bounding box. Afterwards, a uniform distribution defined
on the bounding box area, is used as the \emph{proposal distribution}
for importance sampling. Defining as $q_i$ the uniform distribution
defined on the $i-th$ bounding box, weighted sampling is performed by
drawing samples from $q_i$ and assigning the weights as
in~\eqref{eq:sampling}.

@@

The constrution of the proposal regions is performed in two steps; (a)
solving the optimisation problems as in OMC and (b) constructing the
bounding box regions.

@@

For each set of nuisance variables $\vb_i, i = \{1,2,\ldots,n_1 \}$,
we search for a point $\thetab^*_i$ such that
$d(M_d(\thetab^*_i,\vb_i), \data) < \epsilon$. For notation
convenience, we name the distance $d(M_d(\thetab,\vb_i), \data)$ as
$d_i(\thetab)$.  Obtaining $\theta_i^*$ involves solving the following
optimisation problem:

\begin{subequations}
\begin{alignat}{2}
  &\!\min_{\thetab}        &\qquad& d_i(\thetab) = d(M_d(\thetab^*,\vb_i), \data)\label{eq:optProb}\\
  &\text{subject to} & & d_i(\thetab) \leq \epsilon
\end{alignat}
\end{subequations}
%
The optimisation problem can be treated as unconstrained, accepting
the optimal point $\thetab_i^* = \text{argmin}_{\thetab} d_i(\thetab)$
only if $d_i(\thetab_i^*) < \epsilon$. The nature of the generative
model $M_r(\thetab)$, specifies the properties of the objective
function $d_i$. If $d_i$ is differentiable any gradient-based
optimizer can be used for~\ref{eq:optProb}. The gradients
$\nabla_{\thetab} d_i$ can be either provided in closed form or
approximated by finite differences. In case $d_i$ is not
differentiable, Bayesian Optimisation~\cite{Shahriari2016} provides an
alternative choice. In this scenario, apart from obtaining an optimal
$\thetab_i^* $, a surrogate model $\hat{d}_i$ of the distance $d_i$ is
also fitted; $\hat{d}_i$ can then substitute the actual model in the
following steps\footnote{Such as proposal region construction, sampling etc}.

\subsubsection*{Proposal region estimation}

The acceptance region $\mathcal{S}_i$ is approximated with a bounding
box $\mathcal{\hat{S}}_i$ around it. The bounding box must contain the
part of the acceptance region \emph{around} $\thetab_i^*$; formally
$\{ \thetab : \thetab \in \mathcal{S}_i \quad \text{and} \quad
||\thetab - \thetab_i^*||_2 < M \}$. The condition
$||\thetab - \thetab_i^*||_2 < M$ restricts that if $\mathcal{S}_i$
contains many \emph{disjoint} sets of $\thetab$ that they fulfill
$d_i(\thetab) < \epsilon$, we want the bounding box to match only the
set that contains $\thetab_i^*$. Ideally, we want the bounding box to
be as tight as possible to $\mathcal{S}_i$, to ensure high acceptance
rate in the importance sampling, and big enough for not discarding
valid parts of $\mathcal{S}_i$.

The bounding boxes are built in two steps; firstly, we define their
axes $\mathbf{v}_d$ in the $d$-dimensional space and, then, the limits
along the axes are obtained with a line-search method looking for the
smallest
$\kappa: d_i(\thetab_i^* + \kappa \vb_d) \geq
\epsilon$\footnote{$-\kappa$ is used as well for the opposite
  direction along the search
  line}. Algorithm~\ref{alg:region_construction} describes the method
in-depth. After the bounding boxes construction, a uniform
distribution $q_i$ is defined on each bounding box, and is used as the
proposal region for importance sampling.

### 2021-08-26 latex_files/ch2/2_5_ROMC_algorithmic.tex
small changes

\subsubsection*{Training part}
\noindent
At the training (fitting) part, the goal is the estimation of the
proposal regions $\mathcal{\hat{S}}_i$. The tasks are; (a) sampling
the nuisance variables $\vb_i \sim p(\vb)$ for obtaining the
deterministic functions $d_i(\thetab)$, (b) defining the optimisation
problems $\min_{\thetab} \: g_i(\thetab)$, (c) obtaining
$\thetab_i^*$, $d_i^*$ (d) checking whether $d_i^* \leq \epsilon$ and
(e) computing the proposal distribution $q_i$. If gradients are
available, using a gradient-based method is advised for obtaining
$\thetab_i^*$ much faster. In case $\nabla_{\thetab} g_i$ is available
in closed-form, the procedure is upgraded in terms of both accuracy
and efficiency; If closed-form description is not available, gradients
approximation with finite-differences requires two evaluations of
$g_i$ for \emph{each} parameter $\theta_d$, which can only work in
low-dimensional problems. If $g_i(\thetab)$ is not differentiable, the
Bayesian Optimisation can be used as an alternative solution. In this
scenario, the training part becomes slower due to fitting of the
surrogate model and the blind optimisation steps. After obtaning the
optimal points $\thetab^*$, the training part gets completed by
estimating the proposal regions;
Algorithm~\ref{alg:region_construction} describes the line search
approach for finding the region's boundaries.

>>>>

\subsubsection*{Training part}
\noindent
At the training (fitting) part, the goal is the estimation of the
proposal regions $\mathcal{\hat{S}}_i$. The tasks are; (a) sampling
the nuisance variables $\vb_i \sim p(\vb)$ for obtaining the
deterministic functions $d_i(\thetab)$, (b) defining the optimisation
problems $\min_{\thetab} \: d_i(\thetab)$, (c) obtaining
$\thetab_i^*$, $d_i^*$ (d) checking whether $d_i^* \leq \epsilon$ and
(e) computing the proposal distribution $q_i$. If $d_i(\thetab)$ is
differentiable, using a gradient-based method is advised for obtaining
$\thetab_i^*$ faster. In case $\nabla_{\thetab} d_i$ is available in
closed-form, the procedure is upgraded in terms of both accuracy and
efficiency. If closed-form description is not available, gradients
approximation with finite-differences requires two evaluations of
$d_i$ for \emph{each} parameter $\theta_d$, which can only work in
low-dimensional problems. If $d_i(\thetab)$ is not differentiable, the
Bayesian Optimisation can be used as an alternative solution. In this
scenario, the training part becomes slower due to fitting of the
surrogate model and the blind optimisation steps. After obtaning the
optimal points $\thetab^*$, the training part gets completed by
estimating the proposal regions;
Algorithm~\ref{alg:region_construction} describes the line search
approach for finding the region's boundaries.

### 2021-08-26 latex_files/ch4/4_2_ma2.tex
changed notation to be inline with ROMC paper
small changes

Second-order moving average (MA2) is a probabilistic model commonly
used for univariate time series analysis. The observation at time $t$
is given by:

@@

\begin{center} \label{tab:ma2}
\begin{tabular}{ c|c|c|c|c }
\hline
& $\mu_{\theta_1}$ & $\sigma_{\theta_1}$ & $\mu_{\theta_2}$ & $\sigma_{\theta_2}$ \\
\hline \hline
Rejection ABC & 0.516 & 0.142 & 0.07 & 0.172 \\
\hline
ROMC (gradient-based) & 0.504 & 0.141 & 0.031 & 0.172 \\
\hline
ROMC (Bayesian optimisation) & 0.5 & 0.165 & 0.07 & 0.167 \\
\hline
\end{tabular}
\end{center}

@@

\begin{figure}[ht]
    \begin{center}
        % \input{./latex_files/images/chapter4/ma2_region_3.tex}
        % \input{./latex_files/images/chapter4/ma2_region_3_bo.tex}
        \includegraphics[width=0.49\textwidth]{./latex_files/images/chapter4/ma2_region_3.png}
        \includegraphics[width=0.49\textwidth]{./latex_files/images/chapter4/ma2_region_3_bo.png}
    \end{center}
  \caption[The acceptance region of a specific deterministic simulator.]{The acceptance region in a specific optimisation problem. In the left figure the region obtained with gradient-based optimiser and in the right one with Bayesian Optimisation.}
  \label{fig:ma2_5}
\end{figure}

>>>>

Second-order moving average (MA2) is a probabilistic model commonly
used for time series analysis. The observation at time $t$ is given
by:

@@

\begin{center} \label{tab:ma2}
\begin{tabular}{ c|c|c|c|c }
\hline
& $\mu_{\theta_1}$ & $\sigma_{\theta_1}$ & $\mu_{\theta_2}$ & $\sigma_{\theta_2}$ \\
\hline \hline
Rejection ABC & 0.516 & 0.142 & 0.07 & 0.172 \\
\hline
ROMC (gradient-based) & 0.504 & 0.143 & 0.033 & 0.171 \\
\hline
ROMC (Bayesian optimisation) & 0.502 & 0.166 & 0.079 & 0.168 \\
\hline
\end{tabular}
\end{center}

@@

\begin{figure}[ht]
    \begin{center}
        % \input{./latex_files/images/chapter4/ma2_region_3.tex}
        % \input{./latex_files/images/chapter4/ma2_region_3_bo.tex}
        \includegraphics[width=0.49\textwidth]{./latex_files/images/chapter4/ma2_region_1.png}
        \includegraphics[width=0.49\textwidth]{./latex_files/images/chapter4/ma2_region_1_bo.png}
    \end{center}
  \caption[The acceptance region of a specific deterministic simulator.]{The acceptance region in a specific optimisation problem. In the left figure the region obtained with gradient-based optimiser and in the right one with Bayesian Optimisation.}
  \label{fig:ma2_5}
\end{figure}
